From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/finite.tex
Citations and fixes

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.~\footnote{Note that the indices for the type of permutations
are off-by-one, because we chose $\Fin[n-1]$ to represent generators for permutations on $\Fin[n]$.}

>>>>

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea of this proof is borrowed from~\cite{Molzer-cubical}
~\footnote{Note that the indices for the type of permutations are off-by-one, because we chose $\Fin[n-1]$ to represent
  generators for permutations on $\Fin[n]$.}

### 2021-07-09 Pi+/paper/reversible1.tex
Citations and fixes

These commutative semirig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence to obtain a computational interpretation of the commutative rig
structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
groupoidification of a commutative rig, that is, a symmetric rig groupoid.

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,Carette2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:

>>>>

These commutative rig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence to obtain a computational interpretation of the commutative rig
structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
groupoidification of a commutative rig, that is, a symmetric rig groupoid.

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:

### 2021-07-09 Pi+/paper/ufin.tex
Citations and fixes

Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA}.

@@

\begin{proof}
  (of Prop.~\ref{prop.isFin})
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  applying the first projection.
\end{proof}

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
\footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
  by~\citet{christensenCharacterizationUnivalentFibrations2015}.}

>>>>

Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

@@

\begin{proof}
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  applying the first projection.
\end{proof}

@@

This definition of the groupoid of finite types has also been considered
in~\cite{yorgeyCombinatorialSpeciesLabelled2014}. While $\UFin$ has \emph{all} the finite types, we are also interested
in constructing a subuniverse of finite types of a specified cardinality. To do so, we will start with the subuniverse
$\BAut[T]$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
  by~\citet{christensenCharacterizationUnivalentFibrations2015}.}
